---
title: "Chapter 2"
author: "jim108@gmx.net"
date: "5/14/2020"
output:
  html_notebook:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2E1. 
Probability of rain on Monday?
(2), (4)

## 2E2. 
Pr(Monday|rain)?
(3)


## 2E3.
Probability that it is Monday, given that it is raining? (1), (4)

## 2E4. 
What does it mean to say “the probability of water is 0.7”?

We are following the Laplace-Jeffreys-Cox-Jaynes interpretation of probability. The number of tosses where the globe lands on land devided by the total number of tosses is in expectation 0.7.

## 2M1. 
Compute and plot the grid approximate posterior distribution for
(1) W, W, W
(2) W, W, W, L
(3) L, W, W, L, W, W, W

```{r}
plot_grid_post <- function(k,n, title){
  p_grid <- seq( from=0 , to=1 , length.out=20 )
  prior <- rep( 1 , 20 )
  likelihood <- dbinom( k , size=n , prob=p_grid )
  unstd.posterior <- likelihood * prior
  posterior <- unstd.posterior / sum(unstd.posterior)
  plot( p_grid , posterior , type="b" ,
        xlab="probability of water" , ylab="posterior probability" )
  mtext(title)
}
par(mfrow=c(1,3)) 
plot_grid_post(k=3,n=3, "W, W, W")
plot_grid_post(k=3,n=4, "W, W, W, L")
plot_grid_post(k=5,n=7, "L, W, W, L, W, W, W")
```


## 2M2. 
Assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when p >= 0.5.
```{r}
plot_grid_post_half_prior <- function(k,n, title){
  ## R code 2.3
  # define grid
  p_grid <- seq( from=0 , to=1 , length.out=20 )
  
  # define prior
  prior <- ifelse( p_grid < 0.5, 0, 1)
  
  # compute likelihood at each value in grid
  likelihood <- dbinom( k , size=n , prob=p_grid )
  
  # compute product of likelihood and prior
  unstd.posterior <- likelihood * prior
  
  # standardize the posterior, so it sums to 1
  posterior <- unstd.posterior / sum(unstd.posterior)
  
  plot( p_grid , posterior , type="b" ,
        xlab="probability of water" , ylab="posterior probability" )
  mtext(title)
}
par(mfrow=c(1,3)) 
plot_grid_post_half_prior(k=3,n=3, "W, W, W")
plot_grid_post_half_prior(k=3,n=4, "W, W, W, L")
plot_grid_post_half_prior(k=5,n=7, "L, W, W, L, W, W, W")
```

## 2M3. 
Given: E:="earth", M:="mars",L:="land",
$p(L\mid E) = \frac{3}{10}, \quad p(L\mid M) = 1, \quad p(E) = p(M) = 1/2$.

Show: $p(E\mid L)= 0.23$.

Sol.: $p(E\mid L)=\frac{p(L\mid E)p(E)}{p(L)} = \frac{p(L\mid E)p(E)}{p(L\mid E)p(E) + p(L\mid M)p(M)} = \frac{3}{13}$. 

## 2M4.
Given: cards: B/B, B/W, W/W.

Show: $p(B/B)=\frac{2}{3}$.

Sol.: 

|     | Cards | Ways | Probs |
|-----|-------|------|-------|
|     | B/B   | 2    | 2/3   |
|     | B/W   | 1    | 1/3   |
|     | W/W   | 0    | 0     |
| Sum |       | 3    |       |


## 2M5.
Given: cards: B/B, B/W, W/W, B/B.

Want: p(B/B).

Sol.:

|     | Cards | Ways | Probs |
|-----|-------|------|-------|
|     | B/B   | 4    | 4/5   |
|     | B/W   | 1    | 1/5   |
|     | W/W   | 0    | 0     |
| Sum |       | 5    |       |

## 2M6.
Given: cards: B/B, B/W, W/W, B/B. For every B/B,2 * B/W, 3 * W/W.

Show: p(B/B) = 0.5.

Sol.:

|     | Cards | Ways  | Probs |
|-----|-------|-------|-------|
|     | B/B   | 1 * 2 | 1/2   |
|     | B/W   | 2 * 1 | 1/2   |
|     | W/W   | 0     | 0     |
| Sum |       | 4     |       |

## 2M7.
Given: cards: B/B, B/W, W/W. First draw B/?, second draw W/? without replacement.

Show: p(B/?) = 3/4.

Sol.:

|     | Cards | 1st draw ways | 2nd draw ways | Product | Probs |
|-----|-------|---------------|---------------|---------|-------|
|     | B/B   | 2             | 3             | 6       | 3/4   |
|     | B/W   | 1             | 2             | 2       | 1/4   |
|     | W/W   | 0             | 2             | 0       | 0     |
| Sum |       |               |               | 8       |       |


## 2H1.
Given: Panda bears: A, B. T:="gives birth to twins", S:="gives birth to a single infant".
$p(T\mid A) = \frac{1}{10}$, $p(S\mid A) = \frac{9}{10}$, $p(T\mid B) = \frac{2}{10}$, $p(S\mid B) = \frac{2}{10}$,
$p(A)=p(B)=\frac{1}{2}$.

Want: $p(T\mid T)$.

Sol.: 
$$p(T \mid T) = p(A\mid T) p(T\mid A) + p(B\mid T)p(T\mid B)$$.

$$p(A\mid T) = \frac{p(T\mid A) p(A)}{p(T)} = \frac{1}{3} \quad\Rightarrow\quad p(B\mid T) = \frac{2}{3}.$$
$$\quad p(T) = p(T\mid A) p(A) + p(T\mid B) p(B) = \frac{20}{3}$$
$$\Rightarrow\quad p(T\mid T) = \frac{1}{6}.$$

## 2H2.
Given: same as 2H1.

Want: $p(A\mid T, S)$.

Sol.: 2H1 $\Rightarrow$ $p(A\mid T)=\frac{1}{3}$.


## 2H3.
Given: same as 2H1.

Want: $p(A\mid T)$.

Sol.:
$$p(A \mid T,S) = \frac{p(S,T\mid A) p(A)}{p(A)} = \frac{p(T\mid A) p(S\mid A) p(A)}{p(T\mid A) p(S\mid A)p(A) + p(T\mid B) p(S\mid B) p(B)} =\frac{9}{25}.$$

## 2H4.
Given: 2H1 and $t_A$ := "test indicates panda bear A", $t_B$ := "test indicates panda bear B",
$p(t_A \mid A) = \frac{8}{10}$, $p(t_B \mid A) = \frac{2}{10}$, $p(t_B \mid B) = \frac{15}{20}$, $p(t_A \mid B) = \frac{7}{20}$.

Want: $p(A \mid t_A)$, $p(A\mid t_A, T)$.

Sol.:

$$p(A \mid t_A) = \frac{p(t_A \mid A) p(A)}{p(t_A)} = \frac{16}{23}.$$
$$p(A \mid t_A, T) = \frac{p(t_A,T \mid A) p(A)}{p(t_A,T)} = \frac{p(t_A\mid A) p(T \mid A) p(A)}{p(t_A,T)}= \frac{8}{15}.$$