---
title: "Chapter 11"
author: "jim108@gmx.net"
date: "6/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
```


## 11E1.

Given: Probability $p=0.35$.

Want: Log-odds.

Sol.: 

```{r}
p = 0.35
logit(p)
```


## 11E1.

Given: Log-odds $logodds=3.2$.

Want: Probability.

Sol.: 

```{r}
logodds=3.2
inv_logit(logodds)
```

## 11E3.

Given: Logistic regression with $a=1.7$.

Want: Proportional change of the ods.

Sol.:

```{r}
a<-1.7
exp(a)
```

## 11E4.

Want: Why do Poisson regressions sometimes require the use of an offset? Provide an example.

Sol.:

- Different observations can have different time frames or different distances.
- For example 1.5 manuscrits produced per day at monastery 1 and 3.5 per week at monatery 2.

## 11M1.

Want: Why does likelihood of aggregated binomial data differ from disaggregated binomial data?

Sol.: The likelihood of disaggregated binomial data omits the multiplicity.

## 11M2.

Given: Poisson regression with $a=1.7$.

Want: Change of the outcome.

Sol.:

```{r}
a <- 1.7
exp(a)
```

## 11M3.

Want: Why is the logit link is appropriate for a binomial generalized linear model?

Sol.: Binomial outcomes are between 0 and 1. The logit link restricts the outcome of a linear model in a sensible way (outcome 0 => p=0.5 etc).

## 11M4.

Want: Why is the log link is appropriate for a Poisson generalized linear model?

Sol.: Poisson outcomes are greater than 0. The log link restricts the outcome of the linear model in a sensible way.

## 11M5.

Want: What would it imply to use a logit link for the mean of a Poisson generalized linear model? Example.

Sol.: 

- The rate will be restricted between 0 and 1.
- E.g. restricted growth of plant, which cannot shrink and grow beyound 100 of its original size.

## 11M6.

Want: Constraints for which the binomial and Poisson distributions have maximum entropy. Are the constraints different at all for binomial and Poisson? Why or why not?

Sol.:

- Binomial: Aggregated 0/1 trials with known number of trails.
- Poisson:  Experiments with expected value = variance 
- The constraints are only equal for an infinitly large number of possible outcomes.

##11H1

Given: Chimpanzee data.

Want: Quadratic approximate posterior distribution that includes a unique intercept for each actor. Compare with MCMC.

Sol.:

```{r}
data(chimpanzees)
d <- chimpanzees
d$treatment <- 1 + d$prosoc_left + 2*d$condition

dat_list <- list(
    pulled_left = d$pulled_left,
    actor = d$actor,
    treatment = as.integer(d$treatment) )

l11.4 <- alist(
        pulled_left ~ dbinom( 1 , p ) ,
        logit(p) <- a[actor] + b[treatment] ,
        a[actor] ~ dnorm( 0 , 1.5 ),
        b[treatment] ~ dnorm( 0 , 0.5 )
    )
```


```{r, results = "hide", cache = TRUE}
set.seed(1)
m11.4u <- ulam(l11.4 , data=dat_list , chains=4 , log_lik=TRUE )

precis( m11.4u , depth=2 )
```

```{r}
m11.4q <- quap(l11.4 , data=dat_list)
precis( m11.4q , depth=2 )
```

- There are minor differences in the mean coefficient values, which are all smaller than one standard deviation. 
- The most significant difference is for actor 2. The quadaratic approximation approximates to a normal distribution, which is by definition symmetric. It does not fit to actor 2, who pulls only the right lever.

## 11H2.

Given: Chimpanzee data.

Want: Compare WAIC values of model with intercept for each actor to simpler model.

Sol.:

```{r}
l11.3 <- alist(
        pulled_left ~ dbinom( 1 , p ) ,
        logit(p) <- a + b[treatment] ,
        a ~ dnorm( 0 , 1.5 ),
        b[treatment] ~ dnorm( 0 , 0.5 )
    )
```

```{r}
m11.3q <- quap(l11.3 , data=dat_list)
compare( m11.3q, m11.4q)
```

- The model with actor has a much reliably (dSE<dWAIC) lower WAIC value.

## 11H3.

Given: Eagles data.

Want: (a) Fit with ´ulam´ and ´quap´.

```{r}
library(MASS)
data(eagles)
d <- eagles
levels(d$P)
levels(d$A)
levels(d$V)
```

```{r}
dat_list <- list(
    y = d$y,
    n = d$n,
    P = as.integer(d$P),
    A = as.integer(d$A),
    V = as.integer(d$V))
dat_list
```


```{r}
l11H3 <- alist(
        y ~ dbinom( n , p ),
        logit(p) <- a + bP*P + bV*V + bA*A,
        a ~ dnorm( 0 , 1.5 ),
        c(bP,bV,bA) ~ dnorm( 0 , 1.5 )
)
m11H3q <- quap(l11H3, data=dat_list)

precis( m11H3q , depth=2 )
```

```{r, results = "hide", cache = TRUE}
set.seed(1)
m11H3u <- ulam(l11H3, data=dat_list, chains=4, log_lik = TRUE )

precis( m11H3u , depth=2 )
```

```{r}
compare(m11H3q, m11H3u)
```

- There are minor differences in the mean coefficient values, which are all smaller than one standard deviation. 
- The mcmc approximation has a reliably (dSE<dWAIC) lower WAIC value.

Want: (b) Interpret the estimates. Plot posterior with (1) the predicted probability of success and its 89% interval for each row (i) in the data, as well as (2) the predicted success count and its 89% interval. What different information does each type of posterior prediction provide?

Sol.:

```{r}
set.seed(1)
p <- link(m11H3u)
p.mean <- colMeans(p)
p.PI <- apply(p, 2, PI)

d$p <- d$y / d$n
plot(d$p, col=rangi2, ylab="probability of success", xlab="case", xaxt="n",  ylim = c(0, 1), pch=16)
axis(1, at=1:8, labels=c( "LAL","LAS","LIL","LIS","SAL","SAS","SIL","SIS" ))
points( 1:8 , p.mean )
for(i in 1:8){ lines( c(i, i), p.PI[,i] ) }
```

```{r}
set.seed(1)
y <- sim(m11H3u)
y.mean <- colMeans(y)
y.PI <- apply(y, 2, PI)

plot(d$y, col=rangi2, ylab="predicted success count", xlab="case", xaxt="n", xlim=c(0.75,8.25) , ylim = c(0, 30), pch=16)
axis(1, at=1:8, labels=c( "LAL","LAS","LIL","LIS","SAL","SAS","SIL","SIS" ))
points( 1:8 , y.mean )
for(i in 1:8){ lines(c(i, i), y.PI[,i]) }
```

- According to the plots LAS, LIS, SAS and SIS differ. This could be the case because values at the upper end of the spectrum get squashed by the logit resp. the logistic function. This does not explain the SIS case.

Want: (c) Now try to improve the model. Consider an interaction between the pirate’s size and age (immature or adult). Compare this model to the previous one, using WAIC. Interpret.

Sol.:

```{r, results = "hide", cache = TRUE}
set.seed(1)
l11H3c <- alist(
        y ~ dbinom( n , p ),
        logit(p) <- a + bP*P + bV*V + bA*A + bPA*P*A,
        a ~ dnorm( 0 , 1.5 ),
        c(bP,bV,bA, bPA) ~ dnorm( 0 , 1.5 )
)
m11H3c <- ulam(l11H3c, data=dat_list, chains=4, log_lik = TRUE )
```
```{r}
precis( m11H3c , depth=2 )
```

```{r}
compare(m11H3u, m11H3c)
```

```{r}
set.seed(1)
y <- sim(m11H3c)
y.mean <- colMeans(y)
y.PI <- apply(y, 2, PI)

plot(d$y, col=rangi2, ylab="predicted success count", xlab="case", xaxt="n", xlim=c(0.75,8.25) , ylim = c(0, 30), pch=16)
axis(1, at=1:8, labels=c( "LAL","LAS","LIL","LIS","SAL","SAS","SIL","SIS" ))
points( 1:8 , y.mean )
for(i in 1:8){ lines(c(i, i), y.PI[,i]) }
```

- According to the WAIC the model has inproved.
- The highest successful counts are achieved by a large, adult pirate with a small victim, as expected.
- The lowest successful counts are achieved by a large victims or small and infant pirates, as expected.
- According to the precis output, the interaction effect of pirate size and adult/infant is important.

## 11H4.

Given: Salamander data.

Want: Model the relationship between density and percent cover, using a log-link. Compare ´quap´ vs ´ulam´ approximation. Plot 89 % interval.

Sol.:

```{r}
data("salamanders")
d <- salamanders
head(d)
```

```{r}
dat_list <- list(
    S = d$SALAMAN,
    C = standardize(d$PCTCOVER),
    A = standardize(d$FORESTAGE)
)
head(dat_list)
```

```{r}
l11H4 <-alist(
        S ~ dpois( lambda ),
        log(lambda) <- a + bC*C,
        a ~ dnorm( 0 , 0.5 ),
        bC ~ dnorm( 0 , 0.5 )
        )
m11H4q <- quap(l11H4, data=dat_list)

precis( m11H4q , depth=2 )
```
```{r, results = "hide", cache = TRUE}
set.seed(1)
m11H4u <- ulam(l11H4, data=dat_list, chains=4, log_lik = TRUE)

precis( m11H4u , depth=2 )
```

```{r}
set.seed(1)
compare(m11H4q, m11H4u)
```

```{r}
set.seed(1)

plot( dat_list$C , dat_list$S , xlab="C" , ylab="S" ,
    col=rangi2 , lwd=2 , 
    ylim=c(min(dat_list$S),max(dat_list$S)),
    xlim=c(min(dat_list$C),max(dat_list$C)) )

ns <- 100
C.seq <- seq( from=min(dat_list$C) , to=max(dat_list$C) , length.out=ns )

lambda <- link( m11H4q , data=data.frame( C=C.seq) )
lmu <- apply( lambda , 2 , mean )
lci <- apply( lambda , 2 , PI )
lines( C.seq , lmu , lty=2 , lwd=1.5 )
shade( lci , C.seq , xpd=TRUE )
```

Want: Include ´FORESTAGE´ into the model. Compare. Interpret.

Sol.:

```{r}
l11H4c <-alist(
        S ~ dpois( lambda ),
        log(lambda) <- a + bA*A + bAC*A*C + bC*C,
        a ~ dnorm( 0 , 0.5 ),
        c(bA,bAC,bC) ~ dnorm( 0 , 0.5 )
        )
m11H4cq <- quap(l11H4c, data=dat_list)

precis( m11H4cq , depth=2 )
```

```{r, results = "hide", cache = TRUE}
set.seed(1)
m11H4cu <- ulam(l11H4c, data=dat_list, chains=4, log_lik = TRUE)
```

```{r}
precis( m11H4cu , depth=2 )
```


```{r}
set.seed(1)
compare(m11H4cq, m11H4cu)
```

```{r}
set.seed(1)

par(mfrow=c(1,2))

# vary C
plot( dat_list$C , dat_list$S , xlab="C" , ylab="S" ,
    col=rangi2 , lwd=2 , 
    ylim=c(min(dat_list$S),max(dat_list$S)),
    xlim=c(min(dat_list$C),max(dat_list$C)) )

ns <- 100
C.seq <- seq( from=min(dat_list$C) , to=max(dat_list$C) , length.out=ns )

lambda <- link( m11H4cq , data=data.frame( C=C.seq, A=0) )
lmu <- apply( lambda , 2 , mean )
lci <- apply( lambda , 2 , PI )
lines( C.seq , lmu , lty=2 , lwd=1.5 )
shade( lci , C.seq , xpd=TRUE )

# vary A
plot( dat_list$A , dat_list$S , xlab="A" , ylab="S" ,
    col=rangi2 , lwd=2 , 
    ylim=c(min(dat_list$S),max(dat_list$S)),
    xlim=c(min(dat_list$A),max(dat_list$A)) )

A.seq <- seq( from=min(dat_list$A) , to=max(dat_list$A) , length.out=ns )

lambda <- link( m11H4cq , data=data.frame( A=A.seq, C=0) )
lmu <- apply( lambda , 2 , mean )
lci <- apply( lambda , 2 , PI )
lines( A.seq , lmu , lty=2 , lwd=1.5 )
shade( lci , A.seq , xpd=TRUE )

```

- According to the plots, most datapoints are not in the predicted 89 % interval.
- According to the precis output, the contribution of forestage to the prediction is very small.
- According to the model the number of salamander increases with ground cover. The age of the trees does not matter.