---
title: "Chapter 6"
author: "jim108@gmx.net"
date: "5/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, warning = FALSE}
library(rethinking)
library(dagitty)
```

## 6H1.

Given:

```{r}
dag6M1 <- dagitty( "dag {
    U [unobserved]
    X -> Y
    X <- U <- A -> C -> Y
    U -> B <- C <- V -> Y
}")
coordinates( dag6M1 ) <- list( x=c(A=1,U=0,B=1,C=2,V=3,X=0,Y=2) ,
                               y=c(A=0.5,U=1,B=1.5,C=1,V=1,X=2,Y=2) )
drawdag( dag6M1 )
```

Want: How many paths connect X to Y? Which must be closed? Which variables should you condition on?

Sol.:

```{r}
paths( dag6M1, from="X", to="Y")
adjustmentSets( dag6M1 , exposure="X" , outcome="Y" )
```

## 6H1.

Given: WaffleDivorce data.

Want: DAG, influence waffle houses on divorce rate.

Sol.:

```{r}
library(dagitty)
dag_wd <- dagitty( "dag {
    A -> D
    A -> M -> D
    A <- S -> M
    S -> W -> D
}")
coordinates(dag_wd) <- list( x=c(S=0,W=2,M=1,A=0,D=2) , 
                             y=c(S=0,W=0,M=1,A=2,D=2) )
drawdag( dag_wd )
```

```{r}
data("WaffleDivorce")

d <- WaffleDivorce
head(d)
```

```{r}
d$D <- standardize( d$Divorce )
d$M <- standardize( d$Marriage )
d$A <- standardize( d$MedianAgeMarriage )
d$W <- standardize( d$WaffleHouses )
```

```{r}
mH1 <- quap(
    alist(
        D ~ dnorm( mu , sigma ) ,
        mu <- a[South] + bA*A + bM*M + bW*W,
        a[South] ~ dnorm( 0, 0.5),
        bA[South] ~ dnorm( 0 , 0.5 ) ,
        bM[South] ~ dnorm( 0 , 0.5 ) ,
        bW[South] ~ dnorm( 0 , 0.5 ) ,
        sigma ~ dexp( 1 )
    ) , data = d )

precis( mH1, depth=2)
```

```{r}
bW_south <- coef(mH1)['bW[1]']
bW_not_south <- coef(mH1)['bW[2]']
print(sprintf("Influnce of the number of waffle houses on the the divorce rate in the south: %.2f", bW_south))
print(sprintf("Influnce of the number of waffle houses on the the divorce rate not in the south: %.2f", bW_not_south))
```

## 6H2.

Given: WaffleDivorce data.

Want: Test implied conditional independencies in the model.

Sol.:

```{r}
impliedConditionalIndependencies( dag_wd )
```

Want: (a) Does "A _||_ W | S" hold?


```{r}
dat <- list(
  A = d$A,
  W = d$W,
  S = as.integer(as.factor(d$South))
)
```



```{r}
m6H2a <- quap(
    alist(
        W ~ dnorm( mu , sigma ) ,
        mu <- a[S] + b[S]*A,
        a[S] ~ dnorm( 0 , 1 ) ,
        b[S] ~ dnorm( 0 , 1 ) ,
        sigma ~ dexp( 1 )
    ) , data = dat )

precis( m6H2a, depth = 2)
```

- In the south, age is positively correlated with the number of waffle houses.
- Outside the south, age is not correlated with the number of waffle houses.

Want: (b) Does "D _||_ S | A, M, W" hold?

```{r}
summary(d$A)
```

```{r}
discretize <- function(obj_name, N){
  dcol <- d[,c(obj_name)]
  obj_seq <- seq( from=min(d[dcol>-Inf,c(obj_name)]) , to=max(d[dcol<Inf,c(obj_name)]) , length.out=N-2 )
  return(cut(dcol, breaks=c(-Inf,obj_seq,Inf)))
}
d$A.stage <- discretize("A", 11)
d$W.stage <- discretize("M", 11)
d$M.stage <- discretize("W", 11)
```

```{r}
dat <- list(
  A = as.integer(as.factor(d$A.stage)),
  W = as.integer(as.factor(d$W.stage)),
  S = d$South,
  M = as.integer(as.factor(d$M.stage)),
  D = d$D
)
```

```{r, error=TRUE}
m6H2b <- quap(
    alist(
        D ~ dnorm( mu , sigma ) ,
        mu <- aA[A] + aM[M] + aW[W] + (bA[A] + bM[M] + bW[W])*S,
        aA[A] ~ dnorm( 0 , 1 ) ,
        aM[M] ~ dnorm( 0 , 1 ) ,
        aW[W] ~ dnorm( 0 , 1 ) ,
        bA[A] ~ dnorm( 0 , 1 ) ,
        bM[M] ~ dnorm( 0 , 1 ) ,
        bW[W] ~ dnorm( 0 , 1 ) ,
        sigma ~ dexp( 1 )
    ) , data = dat, start = list(aA=0,aM=0,aW=0,bA=0,bM=0,bW=0) )

precis( m6H2b, depth = 2)
```

- TODO

Want: (c) Does "M _||_ W | S" hold?

```{r}
dat <- list(
  M = d$M,
  W = d$W,
  S = as.integer(as.factor(d$South))
)
```



```{r}
m6H2c <- quap(
    alist(
        W ~ dnorm( mu , sigma ) ,
        mu <- a[S] + b[S]*M,
        a[S] ~ dnorm( 0 , 1 ) ,
        b[S] ~ dnorm( 0 , 1 ) ,
        sigma ~ dexp( 1 )
    ) , data = dat )

precis( m6H2c, depth = 2)
```

- In the south, marriage is weakly positively correlated with the number of waffle houses.
- Outside the south, marriage is not correlated with the number of waffle houses.